---
title: "Swirl Tutorial for dplyr"
author: "Faye Harwell"
date: "September 19, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### One unique aspect of dplyr is that the same set of tools allow you to work with tabular data from a variety of sources, including data frames, data tables, databases and multidimensional arrays. In this lesson, we'll focus on data frames, but everything you learn will apply equally to other formats.

### I've created a variable called path2csv, which contains the full file path to the dataset. Call read.csv() with two arguments, path2csv and stringsAsFactors = FALSE, and save the result in a new variable called mydf. Check ?read.csv if you need help.

### My dataframe is called 'mydf'

### Always remember to load dplyr

```{r}
library(dplyr)
```

### The first step of working with data in dplyr is to load the data into what the package authors call a 'data frame tbl' or 'tbl_df'. Use the following code to create a new tbl_df called cran:

```{r} 
cran <- tbl_df(mydf)
rm("mydf")
cran
```

###  This output is much more informative and compact than what we would get if we printed the original data frame (mydf) to the console.

### First, we are shown the class and dimensions of the dataset. Just below that, we get a preview of the data. Instead of attempting to print the entire dataset, dplyr just shows us the first 10 rows of data and only as many columns as fit neatly in our console. At the bottom, we see the names and classes for any variables that didn't fit on our screen.

### According to the "Introduction to dplyr" vignette written by the package authors, "The dplyr philosophy is to have small functions that each do one thing well." Specifically, dplyr supplies five 'verbs' that cover most fundamental data manipulation tasks: select(), filter(), arrange(), mutate(), and summarize().

### Here is how the select function works!

```{r} 
select(cran,ip_id, package, country)
```

### | The first thing to notice is that we don't have to type cran$ip_id, cran$package, and cran$country, as we normally would when referring to columns of a data frame. The select() function knows we are referring to columns of the cran dataset.

### | Recall that in R, the : operator provides a compact notation for creating a sequence of numbers. For example, try 5:20.

```{r} 
5:20
```

### Normally, this notation is reserved for numbers, but select() allows you to specify a sequence of columns this way, which can save a bunch of typing. Use select(cran, r_arch:country) to select all columns starting from r_arch and ending with country.

```{r} 
#In the dataframe, this is selecting columns r_arch through country.... or at least that is what I think it is doing lol

select(cran, r_arch:country)
```

### Instead of specifying the columns we want to keep, we can also specify the columns we want to throw away. To see how this works, do select(cran, -time) to omit the time column.

```{r} 
select(cran, -time)
```

### Or...

```{r} 
select(cran, -(X:size))
```

### | The == operator asks whether the thing on the left is equal to the thing on the right. If yes, then it returns TRUE. If no, then FALSE. In this case, package is an entire vector (column) of values, so package == "swirl" returns a vector of TRUEs and FALSEs. filter() then returns only the rows of cran corresponding to the TRUEs.

### You can specify as many conditions as you want, separated by commas. For example filter(cran, r_version == "3.1.1", country == "US") will return all rows of cran corresponding to downloads from users in the US running R version 3.1.1. Try it out.

```{r} 
filter(cran, r_version == "3.1.1", country == "US")
```

###  filter(cran, r_version <= "3.0.2", country == "IN") will return all rows for which r_version is less than or equal to "3.0.2" and country is equal to "IN".

```{r} 
filter(cran, r_version <= "3.0.2", country == "IN")
```

### Our last two calls to filter() requested all rows for which some condition AND another condition were TRUE. We can also request rows for which EITHER one condition OR another condition are TRUE. For example, filter(cran, country == "US" | country == "IN") will gives us all rows for which the country variable equals either "US" or "IN". Give it a go.

```{r} 
filter(cran, country == "US" | country == "IN")
```

### | Now, use filter() to fetch all rows for which size is strictly greater than (>) 100500 (no quotes, since size is numeric) AND r_os equals "linux-gnu". Hint: You are passing three arguments to filter(): the name of the dataset, the first condition, and the second condition.

```{r} 
filter(cran, size > 100500, r_os == "linux-gnu")
```

## Want to get rid of NAs in your dataset??? Do this...

### Okay, ready to put all of this together? Use filter() to return all rows of cran for which r_version is NOT NA. Hint: You will need to use !is.na() as part of your second argument to filter().

## THIS IS VERY COOL AND YOU SHOULD PAY ATTENTION

### Now, to order the ROWS of cran2 so that ip_id is in ascending order (from small to large), type arrange(cran2, ip_id). You may want to make your console wide enough so that you can see ip_id, which is the last column.

```{r} 
arrange(cran2,ip_id)
```

### | To do the same, but in descending order, change the second argument to desc(ip_id), where desc() stands for 'descending'. Go ahead.

```{r} 
arrange(cran2, desc(ip_id))
```

### We can also arrange the data according to the values of multiple variables. For example, arrange(cran2, package, ip_id) will first arrange by package names (ascending alphabetically), then by ip_id. This means that if there are multiple rows with the same value for package, they will be sorted by ip_id (ascending numerically). Try arrange(cran2, package, ip_id) now.

```{r} 
arrange(cran2, package, ip_id)
```

###  Arrange cran2 by the following three variables, in this order: country (ascending), r_version (descending), and ip_id (ascending).

```{r} 
arrange(cran2,country,desc(r_version),ip_id)
```

### To illustrate the next major function in dplyr, let's take another subset of our original data. Use select() to grab 3 columns from cran -- ip_id, package, and size (in that order) -- and store the result in a new variable called cran3.

### We want to add a column called size_mb that contains the downloadsize in megabytes. Here's the code to do it:
 
```{r} 
mutate(cran3, size_mb = size / 2^20)
```

### Let's try one more for practice. Pretend we discovered a glitch in the system that provided the original values for the size variable. All of the values in cran3 are 1000 bytes less than they should be. Using cran3, create just one new column called correct_size that contains the correct size.

```{r} 
mutate(cran3, size_mb = size / 2^20, size_gb = size_mb/ 2^10)
```

### | Let's try one more for practice. Pretend we discovered a glitch in the system that provided the original values for the size variable. All of the values in cran3 are 1000 bytes less than they should be. Using cran3, create just one new column called correct_size that contains the correct size.

```{r} 
mutate(cran3,correct_size= size+1000)
```

###  The last of the five core dplyr verbs, summarize(), collapses the dataset to a single row. Let's say we're interested in knowing the average download size. summarize(cran, avg_bytes = mean(size)) will yield the mean value of the size variable. Here we've chosen to label the result 'avg_bytes', but we could have named it anything. Give it a try.

```{r} 
summarize(cran, avg_bytes = mean(size))
```